---
title: "Spotify Song Success Predictions"
author: "Kathleen Phan"
date: "2026-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Introduction

As someone who enjoys listening to many different types of music, I wondered what types of songs are universally considered as "hits". In this project I seek to answer this question, using audio features from Spotify's dataset to build predictive models that classify songs as hits (popularity score ≥ 70) or non-hits.

The dataset includes audio features such as danceability, energy, tempo, and valence. I compare four classification models:

-   Logistic Regression
-   Classification Tree
-   Random Forest
-   LASSO Regression


# 2. Load Libraries and Data

```{r load}
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(glmnet)
library(pROC)

# Load data — update filename if yours is different
spotify <- read_csv("spotify-tracks-dataset.csv")
```


# 3. Data Cleaning and Preparation
```{r clean}
# Select relevant columns
spotify_clean <- spotify %>%
  select(popularity, danceability, energy, loudness, speechiness,
         acousticness, instrumentalness, liveness, valence, tempo,
         explicit, track_genre) %>%
  drop_na()

# Create binary response variable 
spotify_clean <- spotify_clean %>%
  mutate(  # add new 'hit' column 
    hit = factor(ifelse(popularity >= 70, "hit", "not_hit")),  #hit = 1 if popularity >= 70
    explicit = factor(explicit)
  ) %>%
  select(-popularity)  # remove popularity since it defines the response

# Check class balance
table(spotify_clean$hit)
prop.table(table(spotify_clean$hit))
```

# 4. Exploratory Data Analysis

```{r eda of audio features}
# Distribution of key audio features 
spotify_clean %>%
  pivot_longer(cols = c(danceability, energy, valence, tempo),  #reformat columns into rows for ggplot ease
               names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = hit)) +
  geom_density(alpha = 0.5) +  # transparent density curve 
  facet_wrap(~feature, scales = "free") +  # creates separate plots 
  scale_fill_manual(values = c("hit" = "#FFADC7", "not_hit" = "#535353")) +
  labs(title = "Audio Feature Distributions: Hit vs Not Hit",
       x = "Value", y = "Density", fill = "Label") +
  theme_minimal()  #light gridlines, white bkg
```

```{r eda of hits by genre}
# Hit rate by genre (top 15 genres)
spotify_clean %>%
  group_by(track_genre) %>%
  summarise(hit_rate = mean(hit == "hit"), count = n()) %>%
  filter(count >= 100) %>%
  arrange(desc(hit_rate)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(track_genre, hit_rate), y = hit_rate)) +
  geom_col(fill = "#F6C8D1") +
  coord_flip() +  # flip x and y, helps readability
  labs(title = "Hit Rate by Genre (Top 15)",
       x = "Genre", y = "Proportion of Hits") +
  theme_minimal()
```

```{r eda of correlation btwn features}
# Correlation between numeric features
spotify_clean %>%
  select(where(is.numeric)) %>%
  cor() %>%
  round(2)

library(corrplot)
spotify_clean %>%
  select(where(is.numeric)) %>%
  cor() %>%
  corrplot(method = "color", type = "upper", tl.col = "black")
```



# 5. Train/Test Split
```{r split}
set.seed(123)

train_index <- createDataPartition(spotify_clean$hit, p = 0.8, list = FALSE)  # split data into 80% training 20% test
train_data  <- spotify_clean[train_index, ]
test_data   <- spotify_clean[-train_index, ]
```


# 6.1 Logistic Regression
```{r logistic}
log_model <- glm(hit ~ . - track_genre, 
                 data = train_data, 
                 family = binomial)
summary(log_model)

# Predictions
log_probs <- predict(log_model, test_data, type = "response")  # assigns a probability to each set of predictors based on the logistic model
log_preds <- factor(ifelse(log_probs > 0.5, "hit", "not_hit"))  # matchs prob to corresponding response

# Confusion matrix
confusionMatrix(log_preds, test_data$hit, positive = "hit")
```

# 6.2 Classification Tree
```{r tree}
tree_model <- rpart(hit ~ . - track_genre, 
                    data = train_data, 
                    method = "class")

# Plot tree
rpart.plot(tree_model, main = "Classification Tree: Hit Song Prediction")

# Predictions
tree_preds <- predict(tree_model, test_data, type = "class")  # response predictions

# Confusion matrix
confusionMatrix(tree_preds, test_data$hit, positive = "hit")
```

```{r balanced tree}
set.seed(123)
train_balanced <- downSample(x = train_data %>% select(-hit),  #randomly remove rows from 'hit' until same size as 'not hit'
                             y = train_data$hit,
                             yname = "hit")

# Check balance
table(train_data$hit)
table(train_balanced$hit)

# Refit tree using balanced data
tree_model <- rpart(hit ~ . - track_genre,
                    data = train_balanced,
                    method = "class")

rpart.plot(tree_model, main = "Classification Tree: Hit Song Prediction")
```

# 6.3 Random Forest
```{r rf}
set.seed(123)
rf_model <- randomForest(hit ~ . - track_genre,
                         data = train_data,
                         ntree = 100,
                         importance = TRUE)

# Variable importance
varImpPlot(rf_model, main = "Random Forest: Variable Importance")

# Predictions
rf_preds <- predict(rf_model, test_data)
rf_probs <- predict(rf_model, test_data, type = "prob")[, "hit"]

# Confusion matrix
confusionMatrix(rf_preds, test_data$hit, positive = "hit")
```

# 6.4 LASSO Regression
```{r lasso}
# Prepare matrix format for glmnet
x_train <- model.matrix(hit ~ . - track_genre, train_data)[, -1]
y_train <- ifelse(train_data$hit == "hit", 1, 0)

x_test  <- model.matrix(hit ~ . - track_genre, test_data)[, -1]

# Cross-validated LASSO
set.seed(123)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, family = "binomial")

# Best lambda
plot(lasso_cv)
bestlam=lasso_cv$lambda.min #choose lowest lambda
bestlam

# Predictions
lasso_probs <- predict(lasso_cv, x_test, s = "lambda.min", type = "response")
lasso_preds <- factor(ifelse(lasso_probs > 0.5, "hit", "not_hit"))

# Confusion matrix
confusionMatrix(lasso_preds, test_data$hit, positive = "hit")
```

# 7. Model Comparison
```{r compare}
# Function to extract metrics
get_metrics <- function(preds, actual) {
  cm <- confusionMatrix(preds, actual, positive = "hit")
  data.frame(
    Accuracy    = round(cm$overall["Accuracy"], 3),
    Sensitivity = round(cm$byClass["Sensitivity"], 3),
    Specificity = round(cm$byClass["Specificity"], 3),
    Precision   = round(cm$byClass["Pos Pred Value"], 3)
  )
}

# Combine all models
results <- data.frame(
  Model = c("Logistic", "Classification Tree", "Random Forest", "LASSO"),
  rbind(
    get_metrics(log_preds,   test_data$hit),
    get_metrics(tree_preds,  test_data$hit),
    get_metrics(rf_preds,    test_data$hit),
    get_metrics(lasso_preds, test_data$hit)
  )
)

rownames(results) <- NULL  # construct data frame
results
```

```{r compare roc}
# ROC curves
roc_log   <- roc(test_data$hit, log_probs)
roc_rf    <- roc(test_data$hit, rf_probs)
roc_lasso <- roc(test_data$hit, as.vector(lasso_probs))

plot(roc_log,   col = "#FFB3FA", lwd=4, main = "ROC Curves: Model Comparison")
plot(roc_rf,    col = "#191414", add = TRUE)
plot(roc_lasso, col = "#535353", add = TRUE, lty = 2)
legend("bottomright",
       legend = c(paste("Logistic"),
                  paste("Random Forest"),
                  paste("LASSO")),
       col = c("#FFB3FA", "#191414", "#535353"),
       lwd = 2)
```

------------------------------------------------------------------------

# 8. Key findings

-   Random Forest outperformed the other models. It produced more realistic, balanced scores for accuracy, sensitivity, specificity, and precision. 
-   The least important predictors were valence and tempo. 
-   Unsurprisingly, many popular songs tend to have high danceability and energy. 

**Limitations:**

-   "Hit" is defined by an arbitrary popularity threshold of 70. 
-   The dataset may not reflect current trends, nor a wide range of different user preferences. This project investigates a very general, Western audience. 
-   The class imbalance of disproportionately more non-hits than hits affected model performance. This explains why models such as classification trees and lasso has sensitivity scores of 0. These models over-predicted non-hits. 

**Future work:**

-   Investigate difference audiences from difference countries 
-   Try genre-specific models to see if what makes a hit differs by genre


